Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-02 ind_files/01_Introduction.tex
Minor phrasing changes

In population genetics, the relevant high-level forces are the standard evolutionary forces of natural selection, genetic drift, dispersal, and mutation. The description of evolution in terms of these forces was first laid out in formal mathematical terms during the Modern Synthesis by authors such as Wright, Fisher, and Haldane \citep{ewens_mathematical_2004}. The mathematization of the evolutionary forces proved extremely successful, unifying two major schools of thought - Mendelian genetics and Darwinian evolution - that were, at the time, thought to be incompatible \hl{cite something here}.\\
Classical population genetics regarded forces like selection and mutation as fundamental, and the success of this approach during the Modern Synthesis illustrates the value of formulating high-level, abstract models that only provide a `high-level' description of the fundamental processes required to capture the essence of a biological pattern. However, the drama of evolution famously occurs in the ecological theatre \citep{hutchinson_ecological_1965}, and quantities like fitness are not truly fundamental, but instead emerge as the result of various ecological interactions, tradeoffs, and constraints, a fact that can have important consequences for evolution \citep{coulson_putting_2006, kokko_can_2017}. Trying to model such `eco-evolutionary dynamics' has sprouted a rich body of theoretical literature and led to the development of theoretical frameworks like evolutionary game theory and adaptive dynamics, fields which have greatly enriched our understanding of biological populations \citep{brown_why_2016}. Eco-evolutionary population dynamics can broadly be organized under a single unifying framework, the Price equation, that yields all the relevant formal structures under various limits \citep{page_unifying_2002, lion_theoretical_2018}. The Price equation partitions changes in population composition into multiple terms, each of which lends itself to a straightforward interpretation in terms of the high-level evolutionary forces of selection, mutation, and drift, thus providing a useful conceptual framework for thinking about how populations change over time \citep{frank_natural_2012}. The Price equation also leads to a small number of simple yet insightful `fundamental theorems' of population biology \citep{queller_fundamental_2017, lion_theoretical_2018, lehtonen_price_2018} and unifies several various seemingly disjoint formal structures under a single theoretical banner \citep{ lehtonen_price_2020, luque_mirror_2021}\\

@@

Part \ref{part_summary} summarizes the major results of this formalism and presents some simple equations that can be argued to be `fundamental theorems' of population biology in the sense of \cite{queller_fundamental_2017}. These theorems reduce to well-known results such as the Price equation, the replicator-mutator equation from evolutionary game theory, and Fisher's fundamental theorem from population genetics in the infinite population limit. For finite populations, these same theorems predict a new evolutionary force, `noise-induced selection', that has still not found its way into the standard formal canon of evolutionary biology such as the Price equation, and whose significance is only recently being recognized \citep{constable_demographic_2016,mcleod_social_2019,mazzolini_universality_2022, kuosmanen_turnover_2022}. Implications of noise-induced selection are also discussed in part \ref{part_summary}. If one does not care for mathematical details (not advised!), they can skip part \ref{part_theory} entirely and directly read part \ref{part_summary} for the major takeaways of this thesis.

>>>>

In population genetics, the relevant high-level forces are the standard evolutionary forces of natural selection, genetic drift, dispersal, and mutation. The description of evolution in terms of these forces was first laid out in formal mathematical terms during the Modern Synthesis by authors such as Wright, Fisher, and Haldane, an extremely successful venture that unified two major schools of thought --- Mendelian genetics and Darwinian evolution --- that were, at the time, thought to be incompatible \citep{provine_origins_2001}.\\
Classical population genetics regarded forces like selection and mutation as fundamental, and the success of this approach during the Modern Synthesis illustrates the value of formulating high-level, abstract models that only provide a `high-level' description of the fundamental processes required to capture the essence of a biological pattern. However, the evolutionary play famously unfolds in the ecological theatre \citep{hutchinson_ecological_1965}. Thus, quantities like fitness are not truly fundamental but instead emerge as the net result of various ecological interactions, tradeoffs, and constraints, a fact that can have important consequences for evolution \citep{coulson_putting_2006, kokko_can_2017}. Trying to model such `eco-evolutionary dynamics' has sprouted a rich body of theoretical literature and led to the development of theoretical frameworks like evolutionary game theory and adaptive dynamics, fields which have greatly enriched our understanding of biological populations \citep{brown_why_2016}. Eco-evolutionary population dynamics can broadly be organized under a single unifying framework, the Price equation, that yields all the relevant formal structures under various limits \citep{page_unifying_2002, lion_theoretical_2018}. The Price equation partitions changes in population composition into multiple terms, each of which lends itself to a straightforward interpretation in terms of the high-level evolutionary forces of selection, mutation, and drift, thus providing a useful conceptual framework for thinking about how populations change over time \citep{frank_natural_2012}. The Price equation also leads to a small number of simple yet insightful `fundamental theorems' of population biology \citep{queller_fundamental_2017, lion_theoretical_2018, lehtonen_price_2018} and unifies several various seemingly disjoint formal structures under a single theoretical banner \citep{ lehtonen_price_2020, luque_mirror_2021}\\

@@

Part \ref{part_summary} summarizes the major results of this formalism and presents some simple equations that can be argued to be `fundamental theorems' of population biology in the sense of \cite{queller_fundamental_2017}. These theorems reduce to well-known results such as the Price equation, the replicator-mutator equation from evolutionary game theory, and Fisher's fundamental theorem from population genetics in the infinite population limit. For finite populations, these same theorems predict a new evolutionary force, `noise-induced selection', that has still not found its way into the standard formal canon of evolutionary biology such as the Price equation, and whose significance is only recently being recognized \citep{constable_demographic_2016,mcleod_social_2019,mazzolini_universality_2022, kuosmanen_turnover_2022}. Implications of noise-induced selection are also discussed in part \ref{part_summary}. Readers who are averse to or do not care for mathematical details can safely skip part \ref{part_theory} entirely and directly read part \ref{part_summary} for the major takeaways of this thesis, though I strongly recommend working through the formalism properly if possible.
